---
layout: post
title: "Brief MLM tutorial"
date: 2016-09-20 15:30:00 -0700
categories: R tutorials
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { 
    TagSide: "left",
    equationNumbers: {
      autoNumber: "AMS"
    } 
  },
  "HTML-CSS": {
    linebreaks: {
      width: "80% containter"
    }
  }
});
</script>

**Updated: [plotting estimates](#plotting-model-estimates) from real data using `predict`.**

This is a brief description of how to think about multi-level models, especially the link between the formal expression of these models and how that looks in `lme4`. The data set consists of 4 observations per day over 2 days of cortisol in a developmental sample. The relevant theoretical questions are: what individual-level predictors are related to cortisol intercepts and slopes.

<!--more-->

$$
\begin{equation} 
\begin{split}
\text{cort}_{ij} &=\beta_{0j}+\beta_{1j} \cdot \text{time}_{ij}+\epsilon_{ij}\\
\beta_{0j} &=\gamma_{00}+\gamma_{0[\ldots]}\cdot\text{intercept predictors}+u_{0j}\\
\beta_{1j} &=\gamma_{10}+\gamma_{1[\ldots]}\cdot\text{slope predictors}+u_{1j}
\end{split}
\label{corteq}
\end{equation} 
$$

The model in equation \\ref{corteq} says that every *i*th cortisol observation, for the *j*th participant, is predicted by an intercept ($\beta_{0j}$), and a linear lope ($\beta_{1j}$). This is equivalent to a linear model that you're used to in lm: `y ~ 1 + time`. "$\text{time}_{ij}$", in this case, is whatever you want to use -- it's probably going to be the time of day the measurement was taken for the *i*th cortisol observation, for the *j*th participant. 

You're grouping all your observations by subject, and so you can get a random effect (variation across subjects) for both the intercept and slope. That is, each of those parameters gets its own equation, which you see as the second two parts of eq \\ref{corteq}. And just as $\epsilon_{ij}$ is the deviation of every observation from the prediction, $u_{0j}$ and $u_{1j}$ is the deviation of every subject's predicted intercept and slope from the overall mean intercept and slope across all subjects. It just stands in for the fact that we're letting the slope and intercept be different for each person.

For $\beta_{0j}$ and $\beta_{1j}$ you might have multiple predictors you're interested in, so I've just indicated that as "$\gamma_{0/1[\ldots]}\cdot\text{intercept/slope predictors}$". But let's say you have just one subject level predictor for each of those subject-specific slopes and intercepts. I'll use SES as the intercept and slope predictor (though you could, in principle, use different predictors for each parameter). The three equations would be:

$$
\begin{equation} 
\begin{split}
\text{cort}_{ij} &=\beta_{0j}+\beta_{1j} \cdot \text{time}_{ij}+\epsilon_{ij}\\
\beta_{0j} &=\gamma_{00}+\gamma_{01}\cdot\text{SES}_{j}+u_{0j}\\
\beta_{1j} &=\gamma_{10}+\gamma_{11}\cdot\text{SES}_{j}+u_{1j}
\end{split}
\label{corteqfull}
\end{equation}
$$

When we use `lmer` to estimate the model, we only give it one equation. So what we need to do is substitute the equations for $\beta_{0j}$ and $\beta_{1j}$ in to get everything in term of the $\gamma$ parameters (notice that our $\text{time}\_{ij}$ variable gets multiplied through the equation for $\beta_{1j}$ and creates the interaction term $\text{SES}_{j}\times \text{time}$):

$$
\begin{equation} 
\begin{split}
\text{cort}_{ij} =& \gamma_{00}+\gamma_{01}\cdot\text{SES}_{j}+u_{0j}+\\
&(\gamma_{10}+\gamma_{11}\cdot\text{SES}_{j}+u_{1j}) \cdot \text{time}_{ij}+\epsilon_{ij}\\
\text{cort}_{ij} =& \gamma_{00}+\gamma_{01}\cdot\text{SES}_{j}+u_{0j}+\\
&\gamma_{10}\cdot \text{time}_{ij}+\gamma_{11}\cdot\text{SES}_{j}\times \text{time}_{ij}+u_{1j}\cdot \text{time}_{ij} +\epsilon_{ij}\\
\text{cort}_{ij} =& \gamma_{00}+\gamma_{01}\cdot\text{SES}_{j}+\gamma_{10}\cdot \text{time}_{ij}+\gamma_{11}\cdot\text{SES}_{j}\times \text{time}_{ij}+\\
&u_{0j}+u_{1j}\cdot \text{time}_{ij} +\epsilon_{ij}
\end{split}
\label{corteqsingle}
\end{equation}
$$

The final equation in \\ref{corteqsingle} is just reordered so that we group our fixed effects together, and our random effects and error together. Also, I'm just using "$\cdot$" and "$\times$" to set apart parameters and variable interactions a bit.

So now we have our full model equation, and we can give it to `lmer`:

```{r eval=F}
aModel <- lmer(cort ~ 1 + SES + time + SES:time + 
                     (1 + time | SID), 
               data=yourData)
summary(aModel)
```

It doesn't really matter that you have two days of data per person, because we just care about the time of day. You could potentially add another grouping by day, but it might make things more complicated than necessary.

When you get output from `summary`, you'll look at the term `SES:time` to see if a subject's SES predicts their cortisol slope. This is because in equation \\ref{corteqfull}, $\gamma_{11}$ is the parameter for SES predicting $\beta_{1j}$, which is your cortisol slope parameter. This parameter sticks around in the final equation \\ref{corteqsingle} for the interaction term.


## Plotting model estimates

Let's consider a model and some real data: 

```r
lmer(cort ~ 1 + time*ageyrs*SUBTYPE + gender  +steroid + meds +
            (1 | IDENT_SUBID/index4),
     data=aDF)
```

which expands to (grouping terms by the observation-level parameter they seek to explain):

```r
lmer(cort ~ 1 + ageyrs + SUBTYPE + ageyrs:SUBTYPE + gender + steroid + meds +
            time + time:ageyrs + time:SUBTYPE + time:ageyrs:SUBTYPE +
            (1 | IDENT_SUBID/index4),
     data=aDF)
```

```{r}
aDF <- read.csv('./cort_mini.csv')
head(aDF)
```

We have cortisol measured at 4 different times per day (`index4`), over two days, nested within subjects. As I understand it, grouping by measurement within subject helps account for the possibility that some subjects wake up later than others, and so the intercept of the cortisol measurements for each timepoint may deviate from the expected value. 

We have an observation-level variables: _time_; and several subject-level variables: _ageyrs_, _subtype_, _gender_, _steroid_, and _meds_. I think that this `lmer` call implies the following equations (but I don't often work with this kind of nesting), where _i_ is the observation index, _j_ is the measurement-within-subject index (`IDENT_SUBID:index4`), and _k_ is the subject index (`IDENT_SUBID`):

$$
\begin{equation} 
\begin{split}
\text{cort}_{ij} = &\beta_{0jk}+\beta_{1jk} \cdot \text{time}_{ij}+\epsilon_{ijk}\\
\beta_{0jk} = &\gamma_{00k}+u_{0jk}\\
\beta_{1jk} = &\gamma_{10k}\\
\gamma_{00k} = &\delta_{000}+\delta_{001} \cdot \text{ageyrs}+\delta_{002} \cdot \text{SUBTYPE}+\delta_{003} \cdot \text{ageyrs:SUBTYPE}+\\
 & \delta_{004} \cdot \text{gender}+\delta_{005} \cdot \text{steroid}+\delta_{006} \cdot \text{meds}+v_{00k}\\
\gamma_{10k} = &\delta_{100}+\delta_{101} \cdot \text{ageyrs}+\delta_{102} \cdot \text{SUBTYPE}+\delta_{103} \cdot \text{ageyrs:SUBTYPE}\\
\end{split}
\label{realcorteq}
\end{equation}
$$

We'll estimate the paramters for this model using the restricted data set in `aDF` (this is not public data, so we just have a subset of participants).

```{r}
library(lme4)
aMod <- lmer(cort ~ 1 + ageyrs + SUBTYPE + ageyrs:SUBTYPE + gender + steroid + meds +
            time + time:ageyrs + time:SUBTYPE + time:ageyrs:SUBTYPE +
            (1 | IDENT_SUBID/index4),
     data=aDF)
summary(aMod)
```

The model object in `aMod` contains enought information that we can create a new data set and use `predict` to get expected values for plotting the interactions we're interested in. We'd like to plot the cortisol slope (x = time, y = cort) for different ages (grouping lines by integer ages, say), and we probably want a different plot for each subtype. We also need a value for every term in the model, so let's get the mean for our "control" variables: 

```{r}
mean_gender = .5 #gender is coded 1 and 0
mean_steroid = .5 #steroid is coded 1 and 0
mean_meds = .5 #ditto
```

We now want to get all combinations of age (say, every second integer between 4 and 18), subtype (0 and 1), and continuous time (5am - 11pm). We can use `expand.grid` to do this handily:

```{r}
predData <- data.frame(expand.grid(ageyrs = seq(4, 18, 2),
                                   SUBTYPE = c(0, 1),
                                   time = seq(5.0, 23.0, .5),
                                   gender = mean_gender,
                                   steroid = mean_steroid,
                                   meds = mean_meds))
head(predData)
```

Now we can call `predict` to get values for `cort` (read `?predict.merMod` for more info). We'll set `re.form = NA` so that it only estimates the fixed effects. We could get expected values of cort for each of our participants, which is fun, but I'll save that for later. 

```{r}
predData$cort <- predict(object=aMod, newdata=predData, re.form=NA)
head(predData)
```

Now, to plot it using ggplot in the form I hinted at above:

```{r fig.width=6, fig.height=4}
library(ggplot2)
ggplot(predData, aes(x=time, y=cort, group=ageyrs, color=ageyrs))+
       geom_line()+
       facet_wrap(~SUBTYPE)
```   

In the plot above you see an interaction with age for SUBTYPE 1, only. There are some other ways to play with this plot, like faceting by age, or letting the color be dictated by an interaction between the two:


```{r fig.width=6, fig.height=8}
library(ggplot2)
ggplot(predData, aes(x=time, y=cort, group=factor(SUBTYPE), color=factor(SUBTYPE)))+
       geom_line()+
       facet_wrap(~ageyrs)
```   

```{r fig.width=6, fig.height=8}
library(ggplot2)
ggplot(predData, aes(x=time, y=cort, group=interaction(ageyrs, SUBTYPE), color=interaction(ageyrs, SUBTYPE)))+
       geom_line()
```   
