---
layout: post
title: 'Plotting using `predict`'
date: 2016-09-12 
---

[fork this on gitlab](https://gitlab.com/jflournoy/misc-r-projects/tree/master/plot_using_predict)

One of my favorite functions in R is `predict`, and so in response to this recent question on SlackRs#plots ("Anyone know how to plot an interaction at a moderators mean, +1SD, and -1SD?") I thought I'd write up a quick demo of it's usefulness.

<!--more-->

Let's assume a simple interaction effect between two continuous variables:

$$y = \beta_{0} + \beta_{1}x_{1} + \beta_{2}x_{2} + \beta_{3}x_{1}x_{2}+\epsilon$$

```{r}
set.seed(92299)
N <- 500
b0 <- 0
b1 <- .3
b2 <- .5
b3 <- -.2

aDF <- within(data.frame(x1=rnorm(N)), 
     {
	     x2 <- rnorm(N)
	     x1x2 <- x1*x2
	     y <- b0+b1*x1+b2*x2+b3*x1x2+rnorm(N,0,1)
     })

head(aDF)
```

Now that we've generated data, we can fit a model:

```{r}
aMod <- lm(y~1+x1*x2, aDF)
summary(aMod)
```

How do we plot the relationship between _y_ and _x1_ at different levels of _x2_? We can use `predict` to get the expected _y_ for every _x1_ at the mean and +/-1 SD of _x2_ by creating a new data frame with those values.

```{r}
minx1 <- min(aDF$x1)
maxx1 <- max(aDF$x1)
steps <- 100 #we want to get y values for 100 values in the real range of x1 
x1_values <- seq(minx1, 
		 maxx1, 
		 length.out=steps)
#We'll use `rep` to repeat these values for every value of x2 we want below...
x2.sd <- sd(aDF$x2)
x2.mean <- mean(aDF$x2)

newData <- data.frame(x1=rep(x1_values, 3), #for each val of x2
		      x2=rep(c(x2.mean-x2.sd,
			       x2.mean,
			       x2.mean+x2.sd), 
			     each=steps),
		      x2_level=rep(c('-1 SD',
				     'Mean',
				     '+1 SD'),
				   each=steps)) #each val of x2 for all x1
head(newData)
```

Now use `predict` to get y values.

```{r}
newData$y <- predict(aMod,newdata = newData) 
head(newData)
```

Let's plot it using ggplot2.

```{r}
library(ggplot2)
ggplot(newData, aes(x=x1, y=y, group=x2_level, color=x2_level))+
	geom_line()
```

