---
title: "Demonstration of threat from heteroskedastic error to multi-level longitudinal models"
author: "John C. Flournoy"
date: "October 12, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
code_dir <- '~/code_new/'
```

In my work thus far, sorry to say, I haven't taken very seriously the threat posed by non-homogeneous residual variance in estimating multi-level growth models. But I know of no theoretical or empirical reason to think that the adolescents that I study should not differ from age to age (in some systematic way) in the extent to which our models leave variance unmodeled (and thus, resideding in $\epsilon$).

In this short demonstration, 


```{r}
get_a_df_3 <- function(hetsked_err = TRUE){
  if(hetsked_err){
    sigmamat <- matrix(c(1,0,0,0,5,0,0,0,1), nrow = 3)
  } else {
    sigmamat <- matrix(c(1,0,0,0,1,0,0,0,1), nrow = 3)
  }
  ep <- as.vector(MASS::mvrnorm(1e2, Sigma = sigmamat, mu = c(0, 0, 0)))
  X <- matrix(c(rep(1, 1e2*3), 
                c(runif(1e2, 9.5, 10.5), 
                  runif(1e2, 12.5, 13.5), 
                  runif(1e2, 15.5, 16.5))), ncol = 2)
  X[,2] <- X[,2] - 13
  y <- unlist(lapply(1:dim(X)[1], function(i){
    y <- as.numeric(ep[i])
    return(y)
  }))
  
  adf <- data.frame(id = rep(1:100, 3), age_c = X[,2], y = y)
  adf$age <- adf$age_c + 13
  adf$age_c2 <- adf$age_c^2
  return(adf)
}

get_model_diff_p <- function(iter, hetsked_err = TRUE){
  adf <- get_a_df_3(hetsked_err = hetsked_err)
  adf$age_group <- round(adf$age_c)
  lin <- lme4::lmer(y ~ 1 + age_c + (1 + age_c | id), data = adf, REML = FALSE)
  quad <- lme4::lmer(y ~ 1 + age_c + age_c2 + (1 + age_c | id), data = adf, REML = FALSE)
  quad_coef_nlme_t <- NA
  mod_comp_nlme_p <- NA
  
  if(hetsked_err){
    lin_nlme <- try(nlme::lme(
      fixed = y ~ 1 + age_c,
      data = adf,
      random = ~ 1 + age_c | id,
      weights = nlme::varIdent(form = ~ 1 | factor(age_group)),
      method = 'ML', 
      control = nlme::lmeControl(opt = 'optim',
                                 optimMethod = 'L-BFGS-B',
                                 maxIter = 1e4,
                                 msMaxIter = 1e4,
                                 msMaxEval = 1e4)))
    quad_nlme <- try(nlme::lme(
      fixed = y ~ 1 + age_c + age_c2,
      data = adf,
      random = ~ 1 + age_c | id,
      weights = nlme::varIdent(form = ~ 1 | factor(age_group)),
      method = 'ML', 
      control = nlme::lmeControl(opt = 'optim',
                                 optimMethod = 'L-BFGS-B',
                                 maxIter = 1e4,
                                 msMaxIter = 1e4,
                                 msMaxEval = 1e4)))

    
    if(!inherits(quad_nlme, 'try-error')){
      quad_coef_nlme_t <- coef(summary(quad_nlme))['age_c2', 't-value']
    } 
    if(!any(unlist(lapply(list(lin_nlme, quad_nlme), 
                          inherits, 
                          what = 'try-error')))){
      mod_comp_nlme_p <- anova(lin_nlme, quad_nlme)[2,'p-value']
    }
  } 
  return(c(quad_coef_t = coef(summary(quad))['age_c2','t value'],
           mod_comp_p = anova(lin, quad)[2,'Pr(>Chisq)'],
           quad_coef_nlme_t = quad_coef_nlme_t,
           mod_comp_nlme_p = mod_comp_nlme_p))
}
```

```{r fig.width=5, fig.height=5}
ggplot2::ggplot(get_a_df_3(), ggplot2::aes(x = age, y = y)) + 
  ggplot2::geom_point(alpha = .8) + 
  ggplot2::geom_line(alpha = .5, ggplot2::aes(group = id)) + 
  ggplot2::geom_smooth(method = 'glm', formula = y ~ poly(x,2), alpha = .9) + 
  ggplot2::theme_minimal() + 
  ggplot2::labs(title = 'Test')

ggplot2::ggplot(get_a_df_3(hetsked_err = FALSE), ggplot2::aes(x = age, y = y)) + 
  ggplot2::geom_point(alpha = .8) + 
  ggplot2::geom_line(alpha = .5, ggplot2::aes(group = id)) + 
  ggplot2::geom_smooth(method = 'glm', formula = y ~ poly(x,2), alpha = .9) + 
  ggplot2::theme_minimal() + 
  ggplot2::labs(title = 'Null')
```

```{r message=F}
hetsked_fn <- file.path(code_dir, 'misc-r-projects/threats_to_mlm/hetsked_threat.RDS')
hetsked_null_fn <- file.path(code_dir, 'misc-r-projects/threats_to_mlm/hetsked_threat_null.RDS')
if(!file.exists(hetsked_fn) || !file.exists(hetsked_null_fn)){
  many_p <- parallel::mclapply(1:(1e3*parallel::detectCores()), 
                               get_model_diff_p, 
                               hetsked_err = TRUE,
                               mc.cores = parallel::detectCores())
  many_p.df <- data.frame(matrix(unlist(many_p), ncol=length(many_p[[1]]), byrow = T))
  names(many_p.df) <- names(many_p[[1]])
  saveRDS(many_p.df, hetsked_fn)
    
  many_p_null <- parallel::mclapply(1:(1e3*parallel::detectCores()), 
                               get_model_diff_p, 
                               hetsked_err = FALSE,
                               mc.cores = parallel::detectCores())
  many_p_null.df <- data.frame(matrix(unlist(many_p_null), ncol=length(many_p[[1]]), byrow = T))
  names(many_p_null.df) <- names(many_p_null[[1]])
  saveRDS(many_p_null.df, hetsked_null_fn)
} else {
  many_p.df <- readRDS(hetsked_fn)
  many_p_null.df <- readRDS(hetsked_null_fn)
}
```

```{r}
many_p.df <- dplyr::bind_rows(test = many_p.df, null = many_p_null.df, .id = 'model')
many_p.df <- dplyr::mutate(many_p.df,
                           quad_coef_p = 2*pt(abs(quad_coef_t), 
                                              df = 100-4, 
                                              lower.tail = FALSE),
                           quad_coef_norm_p = 2*pnorm(abs(quad_coef_t), 
                                                      lower.tail = FALSE),
                           quad_coef_nlme_p = 2*pt(abs(quad_coef_nlme_t), 
                                              df = 100-4, 
                                              lower.tail = FALSE),
                           quad_coef_norm_nlme_p = 2*pnorm(abs(quad_coef_nlme_t), 
                                                      lower.tail = FALSE))
```

```{r}
ggplot2::ggplot(
  many_p.df,
  ggplot2::aes(x = mod_comp_p)) + 
  ggplot2::geom_histogram(bins = 50,
                          alpha = .3,
                          color = '#555555') + 
  ggplot2::scale_x_continuous(breaks = c(0, .1, .5, 1)) + 
  ggplot2::geom_vline(xintercept = .05, color = 'red', size = 1) + 
  ggplot2::theme_minimal() + 
  ggplot2::labs(
    y = paste0('Count (out of ', dim(many_p.df)[1]/2, ' iterations)'),
    x = as.expression(
      bquote(atop(paste(italic('p'), '-value for the model comparison test'), 
                  paste('Proportion p < .05 for null = ',
                        .(round(mean(many_p.df$mod_comp_p[many_p.df$model == 'null'] < .05), 2)),
                        '; test = ', 
                        .(round(mean(many_p.df$mod_comp_p[many_p.df$model == 'test'] < .05), 2))) )))) + 
  ggplot2::facet_wrap(~model)
ggplot2::ggplot(
  many_p.df,
  ggplot2::aes(x = mod_comp_nlme_p)) + 
  ggplot2::geom_histogram(bins = 50,
                          alpha = .3,
                          color = '#555555') + 
  ggplot2::scale_x_continuous(breaks = c(0, .1, .5, 1)) + 
  ggplot2::geom_vline(xintercept = .05, color = 'red', size = 1) + 
  ggplot2::theme_minimal() + 
  ggplot2::labs(
    y = paste0('Count (out of ', dim(many_p.df)[1]/2, ' iterations)'),
    x = as.expression(
      bquote(atop(paste(italic('p'), '-value for the model comparison test'), 
                  atop(paste('Proportion of models converged = ',
                             .(round(mean(!is.na(many_p.df$mod_comp_nlme_p[many_p.df$model == 'test'])), 2))),
                       paste(' Proportion p < .05 = ', 
                             .(round(mean(many_p.df$mod_comp_nlme_p[many_p.df$model == 'test'] < .05,
                                          na.rm = T), 2)))) )))) + 
  ggplot2::labs(title = 'NLME with residuals allowed to vary by age group for test data')
```

```{r}
ggplot2::ggplot(
  many_p.df,
  ggplot2::aes(x = quad_coef_p)) + 
  ggplot2::geom_histogram(bins = 50,
                          alpha = .3,
                          color = '#555555') + 
  ggplot2::scale_x_continuous(breaks = c(0, .1, .5, 1)) + 
  ggplot2::geom_vline(xintercept = .05, color = 'red', size = 1) + 
  ggplot2::theme_minimal() + 
  ggplot2::labs(
    y = paste0('Count (out of ', dim(many_p.df)[1]/2, ' iterations)'),
    x = as.expression(
      bquote(atop(paste(italic('p'), '-value for the quadratic coef test'), 
                  paste('Proportion p < .05 for null = ',
                        .(round(mean(many_p.df$quad_coef_p[many_p.df$model == 'null'] < .05), 2)),
                        '; test = ', 
                        .(round(mean(many_p.df$quad_coef_p[many_p.df$model == 'test'] < .05), 2))) )))) + 
  ggplot2::facet_wrap(~model)
ggplot2::ggplot(
  many_p.df,
  ggplot2::aes(x = quad_coef_nlme_p)) + 
  ggplot2::geom_histogram(bins = 50,
                          alpha = .3,
                          color = '#555555') + 
  ggplot2::scale_x_continuous(breaks = c(0, .1, .5, 1)) + 
  ggplot2::geom_vline(xintercept = .05, color = 'red', size = 1) + 
  ggplot2::theme_minimal() + 
  ggplot2::labs(
    y = paste0('Count (out of ', dim(many_p.df)[1]/2, ' iterations)'),
    x = as.expression(
      bquote(atop(paste(italic('p'), '-value for the quadratic coef test'), 
                  atop(paste('Proportion of models converged = ',
                             .(round(mean(!is.na(many_p.df$quad_coef_nlme_p[many_p.df$model == 'test'])), 2))),
                       paste(' Proportion p < .05 = ', 
                             .(round(mean(many_p.df$quad_coef_nlme_p[many_p.df$model == 'test'] < .05,
                                          na.rm = T), 2)))) )))) + 
  ggplot2::labs(title = 'NLME with residuals allowed to vary by age group for test data')
```


```{r}

